---
title: "tableclothr"
date: 2023-07-13
author: Haniah Kring 
categories: [package, tables, theming]
---
  
# What is tableclothr?

tableclothr is a package which allows you to easily transform your data into static or dynamic tables. The original version was drafted by [Josiah Ringelberg](https://jringelb.github.io), and I have since grown it into a full fletched package.


## Core Functionality

tableclothr integrates two of the most popular table creation packages together: gt and DT.

1. gt follows in the footsteps of ggplot2, aiming to be a powerful, yet straightforward way of creating beatiful tables in R
2. DT is an R interface for the JavaScript library, DataTables. It's power comes from the ability to create html tables that allow you to interactively filter, sort, and search your data

tableclothr has one primary function: `tablecloth()`.

In gt format:

```{r}
#| layout-nrow: 2
library(tableclothr)
library(gt)

tablecloth(gtcars,
           title = 'Cars Summary',
           group_col = 'year',
           rowname_col = 'mfr',
           aggregate_by = 'median',
           format = 'static',
           theme = 'greyscale')

```
In DT format:

```{r}

library(DT)

```
