---
title: "tableclothr"
date: 2023-07-13
author: Haniah Kring 
categories: [package, tables, theming]
code-fold: true
---
  
# What is tableclothr?

[tableclothr](https://github.com/hrkring/tableclothr) is a package which offers a set of flexible wrappers and supporting functions for the creation and manipulation of gt and DT tables. The original version was drafted by [Josiah Ringelberg](https://jringelb.github.io), and I have since grown it into a full fletched package.


## Core Functionality

tableclothr integrates two of the most popular table creation packages together: gt and DT.

1. gt follows in the footsteps of ggplot2, aiming to be a powerful, yet straightforward way of creating beatiful tables in R
2. DT is an R interface for the JavaScript library, DataTables. It's power comes from the ability to create html tables that allow you to interactively filter, sort, and search your data

tableclothr has one primary function: `tablecloth()`.

In gt format:

```{r}
#| label: The tablecloth function
#| message: false
library(dplyr)
library(tableclothr)
library(gt)

gtcars_table <- tablecloth(gtcars,
           title = 'Cars Summary',
           group_col = 'year',
           rowname_col = 'mfr',
           aggregate_by = 'median',
           format = 'static',
           theme = 'greyscale')

gtcars_table

```

You can add color coding using either the default thresholds, or add your own:

```{r}
#| layout-ncol: 2

## Use the default thresholds
add_color_code(gtcars_table, col = "hp")

## Or specify the thresholds yourself
add_color_code(gtcars_table, col = "hp", thresholds = c(400, 600))

```

In DT format:

```{r}
library(DT)

gtcars_table <- tablecloth(gtcars,
           title = 'Cars Dataset',
           group_col = 'year',
           rowname_col = 'mfr',
           format = 'dynamic',
           theme = 'greyscale')

gtcars_table

```
